Finiteness in Cubical Type Theory

### 2020-06-28 sections/finiteness-predicates.tex
finished instances bit for split enumerability

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Def.~\ref{List} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)
\subsection{Instances}
We have now characterised the informativeness of split enumerability (it's the
same as a split surjection), so we should also look at its
\emph{restrictiveness}.
For this we'll first look at some simple types which are clearly finite, and
we'll see how to prove that's the fact using split enumerability.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{alignat}{3}
    & \text{injective}(f) &&\coloneqq \Pi (x, y : A) , f \; x \equiv f \; y \rightarrow x \equiv y \\
    & A \rightarrowtail B &&\coloneqq \Sigma(f : A \rightarrow B) , \text{injective}(f)
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{equation}
    (A \twoheadrightarrow! \; B) \rightarrow (B \rightarrowtail A)
  \end{equation}
\end{lemma}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{equation}
    \frac{
      A \rightarrowtail B \; \; \; \text{Discrete}(B)
    }{
      \text{Discrete}(A)
    }
  \end{equation}
\end{lemma}

>>>>

      Def.~\ref{split-enum-def} (\AgdaDatatype{\ensuremath{\mathcal{E}!}}) \\
      Eqn.~\ref{container-membership} (\AgdaDatatype{\ensuremath{\in}}) \\
      Eqn.~\ref{sp-surj-eqn}  \\
      Def.~\ref{List} (\(\AgdaDatatype{List}\)) \\
      Eqn.~\ref{container-interp}  \\
      Reassociation \\
      Eqn.~\ref{sp-surj-arrow-eqn}
  \end{minipage}
\end{proof}
In the above proof syntax the
\(\AgdaDatatype{\ensuremath{\equiv \langle{} \rangle{} }}\) connects lines which
are definitionally equal, i.e. they are ``obviously'' equal from the type
checker's perspective.
Clearly, only one line isn't a definitional equality: 
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)
\subsection{Instances}
We have now characterised the informativeness of split enumerability (it's the
same as a split surjection), so we should also look at its
\emph{restrictiveness}.
For this we'll first look at some simple types which are clearly finite, and
we'll see how to prove that's the fact using split enumerability.
\begin{lemma}
  \(\bot\), \(\top\), and \(\AgdaDatatype{Bool}\) are split enumerable.
\end{lemma}
\begin{proof}
  These three types are quite obviously finite: we will show only the proof of
  finiteness for \(\AgdaDatatype{Bool}\) here for brevity's sake.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{agdalisting}
\end{proof}

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{alignat}{3}
    & \text{injective}(f) &&\coloneqq \Pi (x, y : A) , f \; x \equiv f \; y \rightarrow x \equiv y \\
    & A \rightarrowtail B &&\coloneqq \Sigma(f : A \rightarrow B) , \text{injective}(f)
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{equation}
    (A \twoheadrightarrow! \; B) \rightarrow (B \rightarrowtail A)
  \end{equation}
\end{lemma}
\begin{lemma} \label{inj-discrete}
  For any type \(A\) which injects into a discrete type \(B\), \(A\) is
  discrete.
  \begin{equation}
    \frac{
      A \rightarrowtail B \; \; \; \text{Discrete}(B)
    }{
      \text{Discrete}(A)
    }
  \end{equation}
\end{lemma}
